---
title: "Project 01: Billboard Summer Hits"
author: "Thomas Spangler"
output: 
  html_document:
    keep_md: true
    toc: true
    toc_float: true
---

### Dataset

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(plotly)
library(viridis)

billboard_hits <- read_csv("../data/all_billboard_summer_hits.csv", col_types = cols()) %>%
  mutate(
    decade = factor(
      paste0(floor(year / 10) * 10, "s"),
      levels = paste0(seq(1950, 2010, by = 10), "s")
    )
  )
```

The data set used is the all_billboard_summer_hits.csv. It contains audio features for Billboard summer hit songs from 1958 to 2017. The data was downloaded from the GitHub data repository provided and saved in the `data/` folder of this project.

```{r}
str(billboard_hits)

nrow(billboard_hits)

ncol(billboard_hits)

names(billboard_hits)
```
The Billboard summer hits dataset contains 600 observations and 22 variables. Each row represents a song and the variables include information about the song as well as other features such as danceability, energy, loudness, speechiness, acousticness, liveness, valence, and tempo.

```{r}
colSums(is.na(billboard_hits))
```
This shows that all variables have no missing values, so the dataset wont need to be cleaned for missing data before creating the charts.

```{r}
billboard_hits %>%
  select(where(is.numeric)) %>%
  summary()
```
The summary table gives an overview of the numeric variables in the dataset. Several audio features, including danceability, energy, and valence, have fairly wide ranges, showing that the songs vary in how upbeat, energetic, and positive they sound. The median values for danceability, energy, and valence are all above 0.60 which suggests that many summer hits were danceable, energetic, and positive. Based on these statistics, the following visualizations will focus on patterns in the energy, danceability, and valence. These variables are useful for understanding how the sound of popular summer songs has changed over time and how different audio features relate to each other.

### Visualization 1

```{r visualization-1-interactive, fig.alt="Interactive line chart showing the average energy of Billboard summer hits from 1958 to 2017."}
yearly_energy <- billboard_hits %>%
  group_by(year) %>%
  summarize(average_energy = mean(energy, na.rm = TRUE), .groups = "drop")

energy_plot <- ggplot(
  yearly_energy,
  aes(
    x = year,
    y = average_energy,
    text = paste(
      "Year:", year,
      "<br>Average energy:", round(average_energy, 3)
    )
  )
) + geom_line(linewidth = 0.8) + geom_point(size = 1.5) +
  labs(
    title = "Average Energy of Billboard Summer Hits Over Time",
    x = "Year",
    y = "Average Energy"
  ) + theme_minimal()

ggplotly(energy_plot, tooltip = "text")
```

This line graph shows how the average energy of Billboard summer hit songs changed from 1958 to 2017. The average energy appears to increase over time, although there are some decreases in certain years. Earlier years, especially from the late 1950s through the 1970s, generally have lower average energy scores, while many years after the 1980s have higher average energy scores. This chart is interactive, so the reader can hover over each year to see the exact average energy value. This gives the reader more detail than a static chart because they can explore specific years instead of only seeing the overall trend.

### Visualization 2


```{r visualization-2}
ggplot(billboard_hits, aes(x = danceability, y = valence)) + geom_point(alpha = 0.6) + geom_smooth(method = "lm", se = FALSE) +
  labs(
    title = "Danceability and Valence of Billboard Summer Hits",
    x = "Danceability",
    y = "Valence"
  )
```

This scatterplot compares the danceability and valence for songs. The upward trend line suggests a positive relationship between the two variables. Songs with higher danceability often have higher valence which means they tend to sound more positive or upbeat. However, since the points are still spread out, danceability does not perfectly predict how positive songs sound.

### Visualization 3


```{r visualization-3}
billboard_hits <- billboard_hits %>%
  mutate(decade = paste0(floor(year / 10) * 10, "s"))

ggplot(billboard_hits, aes(x = decade, y = danceability)) + geom_boxplot() +
  labs(
    title = "Danceability of Billboard Summer Hits by Decade",
    x = "Decade",
    y = "Danceability"
  )

```

This boxplot shows the distribution of danceability scores for Billboard summer hits across different decades. The median danceability appears to increase from the 1950s and 1960s into the 1990s, 2000s, and 2010s. This suggests that more recent summer hits were more danceable than earlier summer hits. The chart also shows variation within each decade, meaning not every song from a decade follows the same pattern.

### Conclusion

Overall, the Billboard summer hits dataset shows that popular summer songs changed over time in many different ways. The first visualization suggests that the average energy of summer hits increased from 1958 to 2017. The second visualization shows a positive relationship between danceability and valence, meaning more danceable songs often sound more positive or upbeat. The third visualization shows that danceability tended to be higher in more recent decades than in earlier decades. Together, these visualizations suggest that Billboard summer hits have generally become more energetic and danceable over time.

### Report Discussion

I planned to create three charts that would explain how the sound of Billboard summer hits changed over time. The first chart I planned was a line graph showing the average energy of songs by year because a line graph is useful for showing trends over time. The second chart I planned was a scatterplot comparing danceability and valence because I wanted to see whether more danceable songs also tended to sound more positive or upbeat. The third chart I planned was a boxplot showing danceability by decade because this made it easier to compare the distribution of scores across different time periods. 

I applied data visualization and design principles by choosing chart types that matched the kind of data being shown. I used a line graph because time is ordered and the main goal was to show change over time. I used a scatterplot for both variables because they are numeric and the goal was to show the relationship between them. I used a boxplot because it shows the median, spread, and variation within each decade. I also used clear titles and axis labels so the reader could understand what each graph was showing without needing extra explanation. For the scatterplot, I added transparency on the points to reduce overlap and added a trend line to make the overall relationship easier to see.


